One-Particle Density Matrix Polarization Susceptibility Tensors

### 2018-07-23 papers/1_dmatpol/paper-r2.tex
Discussion on uniform fields added.

Most of efforts to theoretically treat extended molecular aggregates are rooted in 
energy\hyp{} (or force field\hyp{})based philosophy. However,
in the past decade there has been considerable interest in density\hyp{}based approaches
of a wide scope and aims.\cite{Piquemal.Cisneros.Reinhardt.Gresh.Darden.JCP.2006,
Mandado.Hermida-Ramon.JCTC.2011,
Sun.Chan.ACR.2016,
Hedegard.Reiher.JCTC.2016} 
In particular, the importance of
one\hyp{}electron densities in chemistry is well known and acknowledged in the community,
though still not solved exactly.\cite{Holas.March.PhysRevA.1991}
%that seem to have an important advantage over the standard energetic approach.
With connection to the above, it would be incredibly insightful 
to establish a first\hyp{}principles connection bewteen the one\hyp{}particle
density and the perturbation in the environment through introducing the
\emph{density matrix susceptibility} (DMS) tensors. % because knowing the density distribution
%is equivalent to knowing the state of the molecule.
The main postulate of the present work is that the associated change 
in the one\hyp{}particle density matrix, $\delta {\bf D}$, can be
represented in a product form of the generalized perturbation susceptibility ${\bf B}_i$
and the environment\hyp{}induced perturbation on a molecule, $\delta {\bf P}_i$,
evaluated at certain site $i$.
%and induced dipole moments $\delta {\BM\upmu}_i$ due to electric fields ${\bf E}_i$ 
%at the centroids of the occupied molecular orbitals,
%
\begin{equation*} \label{e:initial-model}
 \delta {\bf D} \approx \sum_i^{\text{Sites}} \text{[DMS of the $i$th site]} 
  \cdot  \delta {\bf P}_i
  %\rightarrow \sum_i^{\rm occ} {\bf B}^{(i)}_{\mu\nu} \cdot {\bf E}_i \;.
\end{equation*}
%
While $\delta {\bf P}_i$ contains critical information 
about the actual distortion of the electronic cloud,
perturbation susceptibility is assumed to be a property of unperturbed 
molecule and is to be precomputed only once. 
Thus, such DMS could be adapted into an efficient effective 
one\hyp{}electron fragment potential, in principle applicable to an arbitrary
level of theory, as long as the one\hyp{}particle densities are available.
It also means that such a technique would dramatically reduce costs of quantum chemistry calculations
becuase the task of performing full computation on a molecular aggregate
would reduce to much simpler matrix operations.
%
%a simple first\hyp{}principles model for prediction 
%of the perturbation\hyp{}induced one\hyp{}electron density
%change of structurally unrelaxed molecule is proposed here.

@@

It seems intuitive that $\delta {\bf D}$ should 
be directly related to the
polarization\hyp{}induced multipole moments because they are
a measure of the distortion of the electron density distribution due to the external electric
field. 
%Unfortunately, such a mathematical problem is very complicated due to its high\hyp{}order tensor nature
%with respect to the unknown $\delta {\bf D}$ and no approximate analytical solutions have been proposed
%up to date. Therefore, in this paper we adopt a somewhat different
%approach that will require \emph{a priori} knowledge from training set of known cases.
Therefore, 
we start our analysis by considering the induced electronic dipole moment 
of a closed-shell molecule, %which is the first non\hyp{}vanishing moment,
%
\begin{equation} \label{e:dmu-exact}
 \delta {\BM{\upmu}}({\bf r}_Q) = 
     2 \Tr{ 
         \left[ 
              \delta {\bf D} \cdot {\mathbb{M}({\bf r}_Q)}
         \right] } \;.
\end{equation}
%
where ${\mathbb{M}({\bf r}_Q)}$ is AO\hyp{}representation
of the dipole moment operator defined with respect to origin at ${\bf r}_Q$,
%
\begin{equation}\label{e:m}
 \left[ {\mathbb{M}({\bf r}_Q)} \right]_{\alpha\beta} = -e\tbraket{\alpha}{{\bf r} - {\bf r}_Q}{\beta} \;.
\end{equation}
%
(In this Work we shall use the special open bold font for matrices of vectors, so that
in this case ${\mathbb{M}}$ can be considered as a $n \times n$ matrix of vectors of length 3).
Note that if the total charge is conserved, the induced moments are independent on the origin.

@@

Now we can obtain the final expression for the density matrix change 
due to external electric field that reproduces the 
induced dipole moment of the molecule.
The change in the LCAO\hyp{}MO coefficient is parameterized as
%
\begin{equation} \label{e:dC-dmatpol}
 \delta C_{\alpha i} = {\bf b}_{\alpha i}^{(10)} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the susceptibility tensor is defined as
%
\begin{equation} \label{e:susceptibility-b}
  b_{\alpha i;w}^{(10)}= \frac{1}{4} \sum_u^{x, y, z} \left[ {\BM{\alpha}}_i \right]_{uw}
   \left[ \left[ {\bf L}_i  \right]^{-1}_{\rm Left} \right]_{u;\alpha}  
\end{equation}
%
for $w=x,y,z$. 
The superscript `$(10)$' denotes the fact that the susceptibility introduced here
is first\hyp{}order with respect to the electric field and zeroth\hyp{}order with respect to
the electric field gradient.
The result from Eq.~\eqref{e:dC-dmatpol} can be now inserted into 
the expression for $\delta {\bf D}$ from Eq.~\eqref{e:dD-2} to finally give
%
\begin{equation}\label{e:final-model.HF}
 \delta D_{\alpha\beta} \approx \sum_i^{\rm occ} {\bf B}^{(i;10)}_{\alpha\beta} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the \emph{density matrix polarization susceptibility tensor} is
%
\begin{multline}  \label{e:susceptibility-B}
 {\bf B}^{(i;10)}_{\alpha\beta} = %\Big\{
                               C_{\alpha i} {\bf b}_{\beta i}^{(10)} + C_{\beta i} {\bf b}_{\alpha i}^{(10)} \\
                                - \sum_\gamma 
                                 \left( 
               D_{\alpha\gamma} C_{\beta i} + D_{\beta\gamma} C_{\alpha i}
                                 \right)
                                           {\bf b}_{\gamma i}^{(10)}
%
%\Big\} 
 \;.
\end{multline}
%
Hence, the three dimensional vector ${\bf B}^{(i;10)}_{\alpha\beta}$ describes the polarization affinity
of the density matrix element associated with the $\alpha$th and $\beta$th orthogonal AO's.
Transforming these vectors to a non\hyp{}orthogonal AO basis is straightforward.
It is emphasized here that all elements of ${\bf B}_{\alpha\beta}^{(i;10)}$
are unique properties of electronically unperturbed molecule.

@@

To determine the optimum set 
$
 {\bf S} = 
\begin{pmatrix}
{\bf S}^{[1]} &
{\bf S}^{[2]} &
{\bf S}^{[3]}
\end{pmatrix}^T
$, we define the loss function $Z$ 
of the three blocks of adjustable parameters
that is subject to the least\hyp{}squares minimization,
%
\begin{equation}\label{e:Z}
 Z[{\bf S}] = \sum_N^{N_{\rm max}} \left( \delta D^{(N)} - \delta \overline{D}^{(N)} \right)^2 \;.
\end{equation}
%
%where $N_{\rm max}$ is the total number of samples. 
%The linear form of Eq.~\eqref{e:final-model.General.Parameters} ensures that the Hessian 
%of $Z$
%is non\hyp{}singular and independent on the parameters. 
The Hessian of $Z$ computed with respect to the parameters is non\hyp{}singular 
and constant.
Therefore, the exact solution for the optimal parameters is given by the Newton equation
%
\begin{equation}\label{e:Newton}
 {\bf S} = -{\bf H}^{-1} \cdot {\bf g} \;,
\end{equation}
%
where ${\bf g}$ and ${\bf H}$ are the gradient vector and the Hessian matrix, respectively.
%
%\begin{equation}\label{e:Newton.S}
% {\bf S} = 
%\begin{pmatrix}
%{\bf S}^{[1]} &
%{\bf S}^{[2]} &
%{\bf S}^{[3]}
%\end{pmatrix}^T
%\end{equation}
%
The explicit forms of the gradient and Hessian are given in the Appendix~\ref{a:blocks}.
Note that the dimensions of parameter space for the block 1, 2 and 3 are all
equal to $3M$.
%unperturbed molecule. This makes them perfect candidates for general effective fragment parameters.

@@

It is instructive to start first from the discussion of polarization under the uniform electric field
condition because, in this case, it is possible to compute the exact DMS according to Eq.~\eqref{e:dD-Taylor} 
directly by numerical differentiation. Here we consider DMS up to second order, i.e.,
%
\begin{subequations}\label{e:exact-DMS}
 \begin{align}
  {\bf B}^{(10)}_{\alpha\beta} &= \frac{\partial D_{\alpha\beta}}{\partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}} \;, \label{e:B10} \\
  {\bf B}^{(20)}_{\alpha\beta} &= \frac{1}{2} 
     \frac{\partial^2 D_{\alpha\beta}}{\partial {\bf F} \otimes \partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}} \;, \label{e:B20}
 \end{align}
\end{subequations}
%
and the associated error analysis is shown in Figure~\ref{f:fig-1}.
\begin{figure}[h]
\includegraphics[width=0.5\textwidth]{data/dmatpol/water/figure1/fig-1.eps}
\caption{\label{f:fig-1} {\bf Performance of the density matrix polarization susceptibility tensors of water molecule
in moderate uniform electric field.} The notation $(R_{\max}0)$ denotes the maximum rank of the electric field
and zero rank of electric field gradient in the expansion from Eq.~\eqref{e:final-model.General}.
Electric field magnitudes in a range from 0 to 0.035 a.u. were tested in 100 random samples with random direction of electric field.
} 
\end{figure}
%
All the models that are linear with respect to the electric field incorrectly predict the polarization energy.
Interestingly, the predicted values are in perfect correlation with the exact estimates with a slope of $\sim$2.
Incorporation of the second\hyp{}order term from Eq.~\eqref{e:B20} 
obtained by numerical differentiation dramatically reduces the errors
leading to quantitative accuracy (with RMSE of 0.029 kcal/mol and $R^2=99.58\%$;
see also red filled dots in Figure~\ref{f:fig-1}).
What is particularly

@@

@@

The \emph{ab initio} DMS model quantitatively reproduces the interaction energy and the dipole moment 
in weak electric field (RMSE of 0.067~kcal/mol and RMSD of 3.4~mD). In stronger electric fields
the predictions become less quantitative, with high RMSE of 4.2~kcal/mol and RMSD of 14.7~mD in strong electric fields.
Induced quadrupole moment is described by the \emph{ab initio} DMS qualitatively well with $R^2$ not exceeding
60\%. Generalized DMS model of the same rank ($R_{\rm max}$=1, $P_{\rm max}$=0) performs slightly worse
for interaction energies but better for induced multipole moments, particularly decreasing the errors of induced quadrupole
calculations almost by a factor of two in strong electric fields, and by a factor of three in weaker fields.
By and large, extending the parameter space improves the accuracy of the generalized DMS model. 
Including DMS of type ${\bf B}^{(20)}$, i.e., that interact with the square of electric field,
leads to improvement in prediction of all quantities evaluated in all electric field strengths considered in this Work. 
Interestingly, interaction energies are somewhat insensitive to addition of the quadrupolar DMS 
(${\bf B}^{(01)}$, interacting with the electric field gradient), in contrast to induced multipole moments
which exhibit noticeable improvement of accuracy. The most accurate generalized DMS model presented here 
($R_{\rm max}$=2, $P_{\rm max}$=1, which corresponds to including ${\bf B}^{(10)}$, ${\bf B}^{(20)}$ and ${\bf B}^{(01)}$)
works quantitatively well in weak and medium electric fields with $R^2$ being systematically 97.5\% and higher,
and with RMSE of 0.8~kcal/mol in moderate fields. 
However, the errors in the interaction energy
in strong fields are 3.4~kcal/mol, which is quite high as compared to the usually acceptable accuracy 
of 1~kcal/mol in routine quantum chemistry calculations. 
Recently, it was found by Fried et~al. by performing molecular dynamics simulations with a classical 
polarizable force field, that the average electric field experienced by water molecule in liquid phase
in room temperature is around 0.014 a.u. ($\sim$70 MV/cm).\cite{Fried.Wang.Boxer.Ren.Pande.JPCB.2013}
This value corresponds to our moderate electric field set of samples (denoted as `M' in Tables~\ref{t:fields}
and \ref{t:results}) for which the
accuracy of the generalized DMS models is good 
(see also error distributions shown in Figure~\ref{f:fig-2}).
Nevertheless, water molecules can experience
stronger electric fields in certain geometries of hydrogen bonded clusters, 
even up to 0.055 a.u. according to study by Reischl et~al.\cite{Reischl.Kofinger.Dellago.MolPhys.2009} 
Therefore, in strong electric fields
the present state of generalized DMS's must be further optimized in order to reach accuracy of at least
1~kcal/mol.

@@

The need of including at least quadratic terms with respect to the electric field becomes clear
already when considering the simplest case with uniform electric field (Figure~\ref{f:fig-1}).
However, using the linear regression model from Eq.~\eqref{e:final-model.General.Parameters} 
to extract such quadratic dependence
yields wrong polarization energies, despite the one\hyp{}electron properties are accurately reproduced
(Figure~\ref{f:fig-2} and Table~\ref{t:results}).
This is probably because the current generalized DMS model is not fully utilizing the variational flexibility
of the loss function in Eq.~\eqref{e:Z} due to the linear parameterization of a non\hyp{}linear susceptibilities 
(Eq.~\eqref{e:final-model.General}). 
For example, the second\hyp{}order susceptibility, ${\bf B}^{(20)}$, is constrained to
form
%
\begin{equation}
 B_{uw}^{(20)} \approx S_u^{[2]} + S_w^{[2]} \;,
\end{equation}
%
%where ${\bf S}$ is the parameter block. 
leading to symmetric matrices.\footnote{Extending the model by adding additional independent parameter block, i.e.,
$B_{uw}^{(20)} \approx S_u^{[2]} + S_w^{[3]}$, leads to singularity in the Hessian matrix. Therefore,
if the linear regression model is considered, the parameter blocks need to be the same.} 
Note however, that not all symmetric matrices
can be factored out in a way shown above which puts significant limits to a linear regression method
currently used.
It is also natural to expect that $B_{uw}^{(20)}$ does not need to be symmetric if it refers to the distributed site(s).
Although such linear parameterization leads to very simple, one\hyp{}step optimization problem 
from Eq.~\eqref{e:Newton} (Hessian is parameter\hyp{}independent), it
is unfortunately insufficient for an accurate description of two\hyp{}electron effects.
On the other hand, constructing more sophisticated models of $B_{uw}^{(20)}$ and $B_{uw}^{(01)}$ 
requires introducing non\hyp{}linearity to the model, which is out of scope of the present contribution.
This shall be studied in detail elsewhere.

@@

In this Work, the \emph{ab initio} model of the one\hyp{}particle density matrix susceptibility tensors
at the HF level was derived for a closed shell molecule in a spatially non\hyp{}uniform
electric field distribution. It was found that the change in the density matrix element
is approximately proportional to the electric field evaluated at the distributed site associated with the 
center of charge of the 
localized molecular orbital. Subsequently, the \emph{ab initio} model served as a basis
for the formulation of the generalized model that is, in principle, not limited to the HF theory,
and takes into account higher\hyp{}order curvatures of the electric field distribution as well.
The generalized model consisted of a set of adjustable parameters defining the density matrix susceptibility tensors
that are able to interact with the electric field, square of electric field and electric field gradients.
The particular density matrix susceptibilities were determined through the least\hyp{}squares 
linear regression with respect to the training enseble of electrostatically perturbed states.

>>>>

Most of efforts to theoretically treat extended molecular aggregates are rooted in 
energy\hyp{} (or force field\hyp{})based philosophy. However,
in the past decade there has been considerable interest in density\hyp{}based approaches
of a wide scope and aims.\cite{Piquemal.Cisneros.Reinhardt.Gresh.Darden.JCP.2006,
Mandado.Hermida-Ramon.JCTC.2011,
Sun.Chan.ACR.2016,
Hedegard.Reiher.JCTC.2016} 
In particular, the importance of
one\hyp{}electron densities in chemistry is well known and acknowledged in the community,
though still not solved exactly.\cite{Holas.March.PhysRevA.1991}
%that seem to have an important advantage over the standard energetic approach.
With connection to the above, it would be incredibly insightful 
to establish a first\hyp{}principles connection bewteen the one\hyp{}particle
density and the perturbation in the environment through introducing the
\emph{density matrix susceptibility} (DMS) tensors. % because knowing the density distribution
%is equivalent to knowing the state of the molecule.
The main postulate of the present work is that the associated change 
in the one\hyp{}particle density matrix, $\delta {\bf D}$, can be
represented in a product form of the generalized perturbation susceptibility ${\bf B}_i$
and the environment\hyp{}induced perturbation on a molecule, $\delta {\bf P}_i$,
evaluated at certain site $i$.
%and induced dipole moments $\delta {\BM\upmu}_i$ due to electric fields ${\bf E}_i$ 
%at the centroids of the occupied molecular orbitals,
%
\begin{equation*} \label{e:initial-model}
% \delta {\bf D} \approx \sum_i^{\text{Sites}} \text{[DMS of the $i$th site]} 
  \delta {\bf D} \approx \sum_i^{\text{Sites}} {\bf B}_i
  \cdot  \delta {\bf P}_i \;.
  %\rightarrow \sum_i^{\rm occ} {\bf B}^{(i)}_{\mu\nu} \cdot {\bf E}_i \;.
\end{equation*}
%
While $\delta {\bf P}_i$ contains critical information 
about the actual distortion of the electronic cloud,
perturbation susceptibility is assumed to be a property of unperturbed 
molecule and is to be precomputed only once. 
Thus, such DMS could be adapted into an efficient effective 
one\hyp{}electron fragment potential, in principle applicable to an arbitrary
level of theory, as long as the one\hyp{}particle densities are available.
It also means that such a technique would dramatically reduce costs of quantum chemistry calculations
becuase the task of performing full computation on a molecular aggregate
would reduce to much simpler matrix operations.
%
%a simple first\hyp{}principles model for prediction 
%of the perturbation\hyp{}induced one\hyp{}electron density
%change of structurally unrelaxed molecule is proposed here.

@@

It seems intuitive that $\delta {\bf D}$ should 
be directly related to the
polarization\hyp{}induced multipole moments because they are
a measure of the distortion of the electron density distribution due to the external electric
field. 
%Unfortunately, such a mathematical problem is very complicated due to its high\hyp{}order tensor nature
%with respect to the unknown $\delta {\bf D}$ and no approximate analytical solutions have been proposed
%up to date. Therefore, in this paper we adopt a somewhat different
%approach that will require \emph{a priori} knowledge from training set of known cases.
Therefore, 
we start our analysis by considering the induced electronic dipole moment 
of a closed-shell molecule, %which is the first non\hyp{}vanishing moment,
%
\begin{equation} \label{e:dmu-exact}
 \delta {\BM{\upmu}}({\bf r}_Q) = 
     2 \Tr{ 
         \left[ 
              \delta {\bf D} \cdot {\mathbb{M}({\bf r}_Q)}
         \right] } \;.
\end{equation}
%
where ${\mathbb{M}({\bf r}_Q)}$ is AO\hyp{}representation
of the dipole moment operator defined with respect to origin at ${\bf r}_Q$,
%
\begin{equation}\label{e:m}
 \left[ {\mathbb{M}({\bf r}_Q)} \right]_{\alpha\beta} = -e\tbraket{\alpha}{{\bf r} - {\bf r}_Q}{\beta} \;.
\end{equation}
%
(In this work we shall use the special open bold font for matrices of vectors, so that
in this case ${\mathbb{M}}$ can be considered as a $n \times n$ matrix of vectors of length 3).
Note that if the total charge is conserved, the induced moments are independent on the origin.

@@

Now we can obtain the final expression for the density matrix change 
due to external electric field that reproduces the 
induced dipole moment of the molecule.
The change in the LCAO\hyp{}MO coefficient is parameterized as
%
\begin{equation} \label{e:dC-dmatpol}
 \delta C_{\alpha i} = {\bf b}_{\alpha i}^{(10)} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the susceptibility tensor is defined as
%
\begin{equation} \label{e:susceptibility-b}
  b_{\alpha i;w}^{(10)}= \frac{1}{4} \sum_u^{x, y, z} \left[ {\BM{\alpha}}_i \right]_{uw}
   \left[ \left[ {\bf L}_i  \right]^{-1}_{\rm Left} \right]_{u;\alpha}  
\end{equation}
%
for $w=x,y,z$. 
The superscript `$(10)$' underlines the fact that the susceptibility introduced here
is first\hyp{}order with respect to the electric field and zeroth\hyp{}order with respect to
the electric field gradient.
The result from Eq.~\eqref{e:dC-dmatpol} can be now inserted into 
the expression for $\delta {\bf D}$ from Eq.~\eqref{e:dD-2} to finally give
%
\begin{equation}\label{e:final-model.HF}
 \delta D_{\alpha\beta} \approx \sum_i^{\rm occ} {\bf B}^{(i;10)}_{\alpha\beta} \cdot {\bf F}({\bf r}_i)  \;,
\end{equation}
%
where the \emph{density matrix polarization susceptibility tensor} is
%
\begin{multline}  \label{e:susceptibility-B}
 {\bf B}^{(i;10)}_{\alpha\beta} = %\Big\{
                               C_{\alpha i} {\bf b}_{\beta i}^{(10)} + C_{\beta i} {\bf b}_{\alpha i}^{(10)} \\
                                - \sum_\gamma 
                                 \left( 
               D_{\alpha\gamma} C_{\beta i} + D_{\beta\gamma} C_{\alpha i}
                                 \right)
                                           {\bf b}_{\gamma i}^{(10)}
%
%\Big\} 
 \;.
\end{multline}
%
Hence, the three dimensional vector ${\bf B}^{(i;10)}_{\alpha\beta}$ describes the polarization affinity
of the density matrix element associated with the $\alpha$th and $\beta$th orthogonal AO's.
Transforming these vectors to a non\hyp{}orthogonal AO basis is straightforward.
It is emphasized here that all elements of ${\bf B}_{\alpha\beta}^{(i;10)}$
are unique properties of electronically unperturbed molecule.

@@

To determine the optimum set 
$
 {\bf S} = 
\begin{pmatrix}
{\bf S}^{[1]} &
{\bf S}^{[2]} &
{\bf S}^{[3]}
\end{pmatrix}^T
$, we define the loss function $Z$ 
of the three blocks of adjustable parameters
that is subject to the least\hyp{}squares minimization,
%
\begin{equation}\label{e:Z}
 Z[{\bf S}] = \sum_N^{N_{\rm max}} \left( \delta D^{(N)} - \delta \overline{D}^{(N)} \right)^2 \;.
\end{equation}
%
%where $N_{\rm max}$ is the total number of samples. 
%The linear form of Eq.~\eqref{e:final-model.General.Parameters} ensures that the Hessian 
%of $Z$
%is non\hyp{}singular and independent on the parameters. 
The Hessian of $Z$ computed with respect to the parameters is non\hyp{}singular
and parameter\hyp{}independent (constant).
Therefore, the exact solution for the optimal parameters is given by the Newton equation
%
\begin{equation}\label{e:Newton}
 {\bf S} = -{\bf H}^{-1} \cdot {\bf g} \;,
\end{equation}
%
where ${\bf g}$ and ${\bf H}$ are the gradient vector and the Hessian matrix, respectively.
%
%\begin{equation}\label{e:Newton.S}
% {\bf S} = 
%\begin{pmatrix}
%{\bf S}^{[1]} &
%{\bf S}^{[2]} &
%{\bf S}^{[3]}
%\end{pmatrix}^T
%\end{equation}
%
The explicit forms of the gradient and Hessian are given in the Appendix~\ref{a:blocks}.
Note that the dimensions of parameter space for the block 1, 2 and 3 are all
equal to $3M$.
%unperturbed molecule. This makes them perfect candidates for general effective fragment parameters.

@@

It is instructive to start first from the discussion of polarization under the uniform electric field
condition because, in this case, it is possible to compute the exact DMS according to Eq.~\eqref{e:dD-Taylor} 
directly by numerical differentiation. Here we consider DMS up to second order, i.e.,
%
\begin{subequations}\label{e:exact-DMS}
 \begin{align}
  {\bf B}^{(10)}_{\alpha\beta} &= \frac{\partial D_{\alpha\beta}}{\partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}} \;, \label{e:B10} \\
  {\bf B}^{(20)}_{\alpha\beta} &= \frac{1}{2} 
     \frac{\partial^2 D_{\alpha\beta}}{\partial {\bf F} \otimes \partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}} \;. \label{e:B20}
 \end{align}
\end{subequations}
%
The associated error distributions in prediction of polarization energies 
are shown in scatter plot in Figure~\ref{f:fig-1}, whereas
statistical performance descriptors of polarization energies and induced multipole moments
are gathered in Table~XXX.
\begin{figure}[h]
\includegraphics[width=0.5\textwidth]{data/dmatpol/water/figure1/fig-1.eps}
\caption{\label{f:fig-1} {\bf Performance of the density matrix polarization susceptibility tensors of water molecule
in moderate uniform electric field.} The notation $(R_{\max}0)$ denotes the maximum rank of the electric field
and zero rank of electric field gradient in the expansion from Eq.~\eqref{e:final-model.General}.
Electric field magnitudes in a range from 0 to 0.035 a.u. were tested in 100 samples with randomly selected 
directions of electric field.
} 
\end{figure}
%
All the tested models that are linear with respect to the electric field 
incorrectly reproduce the polarization energy.
Interestingly, the predicted values are in perfect correlation with the exact estimates with a slope of $\sim$2.
%especially the \emph{ab initio} models from Eq.~\eqref{e:final-model.HF} and Eq.~\eqref{e:dD-Taylor}
Incorporation of the second\hyp{}order term from Eq.~\eqref{e:B20} 
obtained by numerical differentiation dramatically reduces the errors
leading to quantitative accuracy (with RMSE of 0.029 kcal/mol and $R^2=99.58\%$;
see also red filled circles in Figure~\ref{f:fig-1}). On the other hand,
induced dipole and quadrupole moments are reproduced quantitatively well already by
using first\hyp{}order DMS, ${\bf B}^{(10)}$, with $R^2$ coefficients being systematically
greater than 95\%. This observation can be understood by the fact that the induced dipole
and quadrupole moment is approximately proportional to the external electric field, whereas the polarization
energy exhibits quadratic dependence (c.f. Eq.~\eqref{e:dE-dpol}), hence the need of including at least
quadratic terms in the expansion from Eq.~\eqref{e:dD-Taylor}.

Performance of the generalized DMS models based on the linear regression 
is inferior with respect to the \emph{ab initio} approach in terms of calculations of polarization energy.
Although the linear model (blue open circles in Figure~\ref{f:fig-1})
still predicts the correct slope of $\sim$2 at this level of approximation with RMSE of 1.19 kcal/mol,
yet the quadratic model (red open circles)
improves the performance only slightly, yelding RMSE as high as 0.89 kcal/mol
and produces significant spread of points around the correlation line. This striking difference
between the second\hyp{}order \emph{ab initio} and generalized models originates from the limitations
in the current fitting procedure, that is used to approximate non\hyp{}linear susceptibilities (see further discussion).

@@

To study the capability of the presented distributed models to capture 
density matrix polarization in inhomogeneous electric fields, we analyse water molecule
surrounded by point charges.

@@

The \emph{ab initio} DMS model quantitatively reproduces the interaction energy and the dipole moment 
in weak electric field (RMSE of 0.067~kcal/mol and RMSD of 3.4~mD). In stronger electric fields
the predictions become less quantitative, with high RMSE of 4.2~kcal/mol and RMSD of 14.7~mD in strong electric fields.
Induced quadrupole moment is described by the \emph{ab initio} DMS qualitatively well with $R^2$ not exceeding
60\%. Generalized DMS model of the same rank ($R_{\rm max}$=1, $P_{\rm max}$=0) performs slightly worse
for interaction energies but better for induced multipole moments, particularly decreasing the errors of induced quadrupole
calculations almost by a factor of two in strong electric fields, and by a factor of three in weaker fields.
By and large, extending the parameter space improves the accuracy of the generalized DMS model. 
Including DMS of type ${\bf B}^{(20)}$, i.e., that interact with the square of electric field,
leads to improvement in prediction of all quantities evaluated in all electric field strengths considered in this work. 
Interestingly, interaction energies are somewhat insensitive to addition of the quadrupolar DMS 
(${\bf B}^{(01)}$, interacting with the electric field gradient), in contrast to induced multipole moments
which exhibit noticeable improvement of accuracy. The most accurate generalized DMS model presented here 
($R_{\rm max}$=2, $P_{\rm max}$=1, which corresponds to including ${\bf B}^{(10)}$, ${\bf B}^{(20)}$ and ${\bf B}^{(01)}$)
works quantitatively well in weak and medium electric fields with $R^2$ being systematically 97.5\% and higher,
and with RMSE of 0.8~kcal/mol in moderate fields. 
However, the errors in the interaction energy
in strong fields are 3.4~kcal/mol, which is quite high as compared to the usually acceptable accuracy 
of 1~kcal/mol in routine quantum chemistry calculations. 
Recently, it was found by Fried et~al. by performing molecular dynamics simulations with a classical 
polarizable force field, that the average electric field experienced by water molecule in liquid phase
in room temperature is around 0.014 a.u. ($\sim$70 MV/cm).\cite{Fried.Wang.Boxer.Ren.Pande.JPCB.2013}
This value corresponds to our moderate electric field set of samples (denoted as `M' in Tables~\ref{t:fields}
and \ref{t:results}) for which the
accuracy of the generalized DMS models is good 
(see also error distributions shown in Figure~\ref{f:fig-2}).
Nevertheless, water molecules can experience
stronger electric fields in certain geometries of hydrogen bonded clusters, 
even up to 0.055 a.u. according to study by Reischl et~al.\cite{Reischl.Kofinger.Dellago.MolPhys.2009} 
Therefore, in strong electric fields
the present state of generalized DMS's must be further optimized in order to reach accuracy of at least
1~kcal/mol.

@@

The need of including at least quadratic terms with respect to the electric field becomes clear
already when considering the simplest case with uniform electric field (Figure~\ref{f:fig-1}).
However, using the linear regression model from Eq.~\eqref{e:final-model.General.Parameters} 
to extract such quadratic dependence
yields wrong polarization energies, despite the one\hyp{}electron properties are accurately reproduced
(Figure~\ref{f:fig-2} and Table~\ref{t:results}).
This is probably because the current generalized DMS model is not fully utilizing the variational flexibility
of the loss function in Eq.~\eqref{e:Z} due to the linear parameterization of a non\hyp{}linear susceptibilities 
(Eq.~\eqref{e:final-model.General}). 
For example, the second\hyp{}order susceptibility, ${\bf B}^{(20)}$, is constrained to
form
%
\begin{equation}
 B_{uw}^{(20)} \approx S_u^{[2]} + S_w^{[2]} \;,
\end{equation}
%
%where ${\bf S}$ is the parameter block. 
leading to symmetric matrices.\footnote{Extending the model by adding additional independent parameter block, i.e.,
$B_{uw}^{(20)} \approx S_u^{[2]} + S_w^{[3]}$, leads to singularity in the Hessian matrix. Therefore,
if the linear regression model is considered, the parameter blocks need to be the same.} 
Note however, that not all symmetric matrices
can be factored out in a way shown above which puts significant limits to a linear regression method
currently used.
It is also natural to expect that $B_{uw}^{(20)}$ does not need to be symmetric if it refers to the distributed site(s).
Although such linear parameterization enables a relatively simple, one\hyp{}step optimization
from Eq.~\eqref{e:Newton}, 
it is unfortunately insufficient for an accurate description of two\hyp{}electron effects which manifest
here in energy by the Fermi electron correlation at the HF level.
On the other hand, constructing more sophisticated models of $B_{uw}^{(20)}$ and $B_{uw}^{(01)}$ 
requires introducing non\hyp{}linearity to the model, which is out of scope of the present contribution.
This shall be studied in detail elsewhere.

@@

In this work, the \emph{ab initio} model of the one\hyp{}particle density matrix susceptibility tensors
at the HF level was derived for a closed shell molecule in a spatially non\hyp{}uniform
electric field distribution. It was found that the change in the density matrix element
is approximately proportional to the electric field evaluated at the distributed site associated with the 
center of charge of the 
localized molecular orbital. Subsequently, the \emph{ab initio} model served as a basis
for the formulation of the generalized model that is, in principle, not limited to the HF theory,
and takes into account higher\hyp{}order curvatures of the electric field distribution as well.
The generalized model consisted of a set of adjustable parameters defining the density matrix susceptibility tensors
that are able to interact with the electric field, square of electric field and electric field gradients.
The particular density matrix susceptibilities were determined through the least\hyp{}squares 
linear regression with respect to the training enseble of electrostatically perturbed states.
